\huge The human visual system preserves the hierarchy of 2-dimensional pattern regularity

### 2021-06-28 Writeup/final version for upload/kohler_clarke_2021.tex
version for upload created

\documentclass[11pt, twoside]{article}
\usepackage{boisik}
\usepackage[OT1]{fontenc}
\usepackage[affil-it]{authblk}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{multicol}
\usepackage{natbib}
\usepackage{geometry}
\usepackage{amsmath}
\usepackage{lineno}
\usepackage[dvipsnames]{xcolor}
\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{
    colorlinks,
    linkcolor={red!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!80!black}
}

@@

\begin{figure}[hptb]
\centering
\includegraphics[width=0.75\linewidth]{../figures/figure1.pdf}
\caption{Subgroup relationships with indices 2 (solid lines) and 3 (dashed line) are shown in (A). All other relationships can be inferred by identifying the shortest path through the hierarchy, and multiplying the subgroup indices. For example, \textit{P1} is related to \textit{P6} through \textit{P6}\textrightarrow\textit{P3} (index 2) and \textit{P3}\textrightarrow\textit{P1} (index 3) so \textit{P1} is also a subgroup of \textit{P6} with index $3 \times 2 = 6$. We also show enlarged versions of some of the subgroup relationships involving \textit{P6} (B, shown in red) and \textit{PMM} (C, shown in blue) and highlight the symmetries within the subgroups to emphasize how the supergroup can be generated by adding additional transformations to the subgroup. Illustration adapted from \citet{wade_crystal_1993}.}
\label{fig:subgroups}
\end{figure}

\begin{figure}[hp]
\centering
\includegraphics[width=0.75\linewidth]{../figures/figure2.png}
\caption{Examples of each of the 16 wallpaper groups are shown in the left- and right-most column of the figures, next to the corresponding SSVEP (center-left) and psychological (center-right) data from each group. The SSVEP data are odd-harmonic-filtered cycle-average waveforms. In each cycle, a \textit{P1} exemplar was shown for the first 600 ms, followed by the original exemplar for the last 600 ms. Errorbars are standard error of the mean. Psychophysical data are presented as boxplots reflecting the distribution of display duration thresholds. The 16 groups are ordered by the strength of the SSVEP response, to highlight the range of response amplitudes.}
\label{fig:summary}
\end{figure}

The two datasets we present here make it possible to assess the extent to which both behavior and brain responses follow the hierarchy of complexity expressed by the subgroup relationships. Based on previous brain imaging work showing that patterns with more axes of symmetry produce greater activity in visual cortex \citep{Sasaki_2005, Tyler_2005,RN1959, RN1725, keefe_2018}, we hypothesized that more complex groups would produce larger SSVEPs. For the psychophysical data, we hypothesized that more complex groups would lead to shorter symmetry detection thresholds, based on previous data showing that under a fixed presentation time, discriminability increases with the number of symmetry axes in the pattern \citep{wagemans_1991}. Our results confirm both hypotheses, and show that activity in human visual cortex is remarkably consistent with the hierarchical relationships between the wallpaper groups, with SSVEP amplitudes and psychophysical thresholds following these relationships at a level that is far beyond chance. The human visual system thus appears to encode all of the fundamental symmetries using a representational structure that closely approximates the subgroup relationships from group theory.

@@

\begin{figure}[hp]
\centering
\includegraphics[width=0.75\linewidth]{../analysis/plots/subgroup_comp_eeg_rms.png}
\caption{Posterior distributions for the difference in mean SSVEP RMS amplitude. Colour coding of the text indicates the index of the subgroup, while the colour of the filled distribution relates to the conditional probability that the difference in means is greater than zero. We can see that 55/63 subgroup relationships have $p(\Delta <0 | data)>0.99$.}
\label{fig:eeg_comp}
\end{figure}

\begin{figure}[hptb]
\centering
\includegraphics[width=0.75\linewidth]{../analysis/plots/subgroup_comp_psychophysical.png}
\caption{Posterior distributions for the difference in mean symmetry detection threshold durations. Colour coding of the text indicates the index of the subgroup, while the colour of the filled distribution relates to the conditional probability that the difference in means is smaller than zero. We can see that 43/63 subgroup relationships have $p(\Delta >0 | data)>0.99$.}
\label{fig:durations_comp}
\end{figure}

@@

\begin{figure}[tbp]
\centering
\includegraphics[width=0.5\linewidth]{../analysis/plots/model_roc_style.pdf}
\caption{This plot shows the proportion of subgroup relationships that satisfy $p(\Delta >0 | data) > x$ for the SSVEP data and $p(\Delta <0 | data) > x$ for the psychophysical data. We can see that if we take $x = 0.95$ as our threshold, the subgroup relationships are preserved in 56/63 = 89\% and 48/63 = 76\% of the comparisons for the primary SSVEP and threshold duration datasets, receptively. This compares to the 32/63= 51\% and 22/63 = 35\% for the SSVEP control datasets.}
\label{fig:roc_style}
\end{figure}

@@

\subsection*{Bayesian Analysis of SSVEP and Psychophysical data}
Bayesian analysis was carried out using R (v3.6.1) \citep{R} with the \texttt{brms} package (v2.9.0) \citep{burkner2017} and rStan (v2.19.2 \citep{rStan}). The data from each experiment were modelled using a Bayesian generalised mixed effect model with wallpaper group being treated as a 16-level factor, and random effects for participant. The SSVEP data and symmetry detection threshold durations were modelled using log-normal distributions with weakly informative, $ \mathcal{N}(0, 2)$, priors. After fitting the model to the data, samples were drawn from the posterior distribution of the two datasets, for each wallpaper group. These samples were then recombined to calculate the distribution of differences for each of the 63 pairs of subgroup and supergroup. These distributions were then summarised by computing the conditional probability of obtaining a positive (negative) difference, $p(\Delta | \text{data})$. For further technical details, please see the Supplementary Materials where the full R code, model specification, prior and posterior predictive checks, and model diagnostics, can be found.

@@

% Bibliography
\begin{multicols}{2}
\small
\bibliographystyle{apalike} 
\bibliography{literature}
\end{multicols}

>>>>

\documentclass[11pt, twoside]{article}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage[affil-it]{authblk}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{multicol}
\usepackage{natbib}
\usepackage{geometry}
\usepackage{amsmath}
\usepackage{lineno}
\usepackage[dvipsnames]{xcolor}
\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{
    colorlinks,
    linkcolor={red!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!80!black}
}

@@

\begin{figure}[hptb]
\centering
\includegraphics[width=0.75\linewidth]{figure1.pdf}
\caption{Subgroup relationships with indices 2 (solid lines) and 3 (dashed line) are shown in (A). All other relationships can be inferred by identifying the shortest path through the hierarchy, and multiplying the subgroup indices. For example, \textit{P1} is related to \textit{P6} through \textit{P6}\textrightarrow\textit{P3} (index 2) and \textit{P3}\textrightarrow\textit{P1} (index 3) so \textit{P1} is also a subgroup of \textit{P6} with index $3 \times 2 = 6$. We also show enlarged versions of some of the subgroup relationships involving \textit{P6} (B, shown in red) and \textit{PMM} (C, shown in blue) and highlight the symmetries within the subgroups to emphasize how the supergroup can be generated by adding additional transformations to the subgroup. Illustration adapted from \citet{wade_crystal_1993}.}
\label{fig:subgroups}
\end{figure}

\begin{figure}[hp]
\centering
\includegraphics[width=0.75\linewidth]{figure2.pdf}
\caption{Examples of each of the 16 wallpaper groups are shown in the left- and right-most column of the figures, next to the corresponding SSVEP (center-left) and psychological (center-right) data from each group. The SSVEP data are odd-harmonic-filtered cycle-average waveforms. In each cycle, a \textit{P1} exemplar was shown for the first 600 ms, followed by the original exemplar for the last 600 ms. Errorbars are standard error of the mean. Psychophysical data are presented as boxplots reflecting the distribution of display duration thresholds. The 16 groups are ordered by the strength of the SSVEP response, to highlight the range of response amplitudes.}
\label{fig:summary}
\end{figure}

The two datasets we present here (data and analysis code has been made available on \href{https://osf.io/f3ex8/}{OSF}) make it possible to assess the extent to which both behavior and brain responses follow the hierarchy of complexity expressed by the subgroup relationships. Based on previous brain imaging work showing that patterns with more axes of symmetry produce greater activity in visual cortex \citep{Sasaki_2005, Tyler_2005,RN1959, RN1725, keefe_2018}, we hypothesized that more complex groups would produce larger SSVEPs. For the psychophysical data, we hypothesized that more complex groups would lead to shorter symmetry detection thresholds, based on previous data showing that under a fixed presentation time, discriminability increases with the number of symmetry axes in the pattern \citep{wagemans_1991}. Our results confirm both hypotheses, and show that activity in human visual cortex is remarkably consistent with the hierarchical relationships between the wallpaper groups, with SSVEP amplitudes and psychophysical thresholds following these relationships at a level that is far beyond chance. The human visual system thus appears to encode all of the fundamental symmetries using a representational structure that closely approximates the subgroup relationships from group theory.

@@

\begin{figure}[hp]
\centering
\includegraphics[width=0.75\linewidth]{subgroup_comp_eeg_rms.png}
\caption{Posterior distributions for the difference in mean SSVEP RMS amplitude. Colour coding of the text indicates the index of the subgroup, while the colour of the filled distribution relates to the conditional probability that the difference in means is greater than zero. We can see that 55/63 subgroup relationships have $p(\Delta >0 | data)>0.99$.}
\label{fig:eeg_comp}
\end{figure}

\begin{figure}[hptb]
\centering
\includegraphics[width=0.75\linewidth]{subgroup_comp_psychophysical.png}
\caption{Posterior distributions for the difference in mean symmetry detection threshold durations. Colour coding of the text indicates the index of the subgroup, while the colour of the filled distribution relates to the conditional probability that the difference in means is smaller than zero. We can see that 43/63 subgroup relationships have $p(\Delta <0 | data)>0.99$.}
\label{fig:durations_comp}
\end{figure}

@@

\begin{figure}[tbp]
\centering
\includegraphics[width=0.5\linewidth]{model_roc_style.pdf}
\caption{This plot shows the proportion of subgroup relationships that satisfy $p(\Delta >0 | data) > x$ for the SSVEP data and $p(\Delta <0 | data) > x$ for the psychophysical data. We can see that if we take $x = 0.95$ as our threshold, the subgroup relationships are preserved in 56/63 = 89\% and 48/63 = 76\% of the comparisons for the primary SSVEP and threshold duration datasets, receptively. This compares to the 32/63= 51\% and 22/63 = 35\% for the SSVEP control datasets.}
\label{fig:roc_style}
\end{figure}

@@

\subsection*{Bayesian Analysis of SSVEP and Psychophysical data}
Bayesian analysis was carried out using R (v3.6.1) \citep{R} with the \texttt{brms} package (v2.9.0) \citep{burkner2017} and rStan (v2.19.2 \citep{rStan}). The data from each experiment were modelled using a Bayesian generalised mixed effect model with wallpaper group being treated as a 16-level factor, and random effects for participant. The SSVEP data and symmetry detection threshold durations were modelled using log-normal distributions with weakly informative, $ \mathcal{N}(0, 2)$, priors. After fitting the model to the data, samples were drawn from the posterior distribution of the two datasets, for each wallpaper group. These samples were then recombined to calculate the distribution of differences for each of the 63 pairs of subgroup and supergroup. These distributions were then summarised by computing the conditional probability of obtaining a positive (or for the psychophysical data, negative) difference, $p(\Delta | \text{data})$. For further technical details, please see the Supplementary Materials where the full R code, model specification, prior and posterior predictive checks, and model diagnostics, can be found.

@@

% Bibliography
\begin{multicols}{2}
\small
\bibliographystyle{apalike} 
\bibliography{kohler_clarke_2021.bbl}
\end{multicols}
